An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-12-27 article/article.tex
revised until 2.2

An implicit or simulator-based model is a parameterized stochastic
data generating mechanism. The key characteristic of these models is
that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterized
family of probability density functions
\(\{ p(\yb|\thetab)\}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these scenarios, we can only a
access the simulator \( m_r(\thetab) \), i.e., a black-box mechanism
(computer code) that generates samples \(\yb\) in a stochastic manner
from a set of parameters \(\thetab\). We denote the process of
obtaining samples from the simulator with
\( m_r(\thetab) \rightarrow \yb \). It is feasible to isolate the
randomness of the simulator by introducing the nuisance random
variables denoted by \(\ub \sim p(\ub)\). Therefore, for a tuple
\((\thetab, \ub)\), the simulator turns to a deterministic mapping
\(g\), such that \(\yb=\simulator(\thetab,\ub)\). In terms of computer
code, the randomness of a random process is governed by the global
seed. Even though each software may handle the random generation
process in a different way\footnote{For example, at
  \pkg{Numpy}~\citet{harris2020array}, the pseudo-random number
  generation is based on a global state, whereas, in
  \pkg{JAX}~\citet{jax2018github} random functions consume a key that
  is passed as parameter. }, in all cases, by sampling an integer value
and setting it as the initial seed converts the simulation to a
deterministic sequence of steps.

Simulator-based models provide considerable modeling freedom; any
physical process that can be conceptualized as a computer program of
finite steps can be modeled as a simulator-based model without any
compromise. The modeling freedom allows for any amount of hidden
(unobserved) internal variables or rule-based decisions. Hence,
implicit models are often used to model physical phenomena in the
natural sciences such as, e.g., genetics, epidemiology or
neuroscience. Further background on simulator-based models and
example applications can be found in the articles
by~\citet{gutmann2016, lintusaari2017, sisson2018, cranmer2020}.

The modeling freedom of simulator-based models, however, comes at the
price of difficulties in inferring their parameters. Denoting the
observed data as \(\data\), the main difficulty is that the likelihood
function \(l(\thetab) = p(\data|\thetab)\) is generally
intractable. To better see the sources of the intractability, and to
address them, we go back to the basic characterization of the
likelihood as the (rescaled) probability of a parameter of the model
to generate data \(\yb\) that is similar to the observed data
\(\data\). More formally, the likelihood \(l(\thetab)\) equals
\begin{equation} \label{eq:likelihood}
  l(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in b_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around \(\data\)
that is defined via a distance function \(d\), i.e., \ \(\region(\data)
:= \{\yb: d(\yb, \data) \leq \epsilon \}\).

The basic characterization of the likelihood in~\eqref{eq:likelihood}
highlights two sources of intractability; the first is the computation
of the probability
\(\Pr(\simulator(\thetab,\ub) \in \region(\data))\), the second is the
limit of \(\epsilon \to 0\). Approximating the probability with
samples becomes computationally infeasible if \(\epsilon\) is too
small. Hence, a large class of inference methods work with
\(\epsilon > 0\), which leads to the approximate likelihood function
\(l_{d, \epsilon}(\thetab)\)
\begin{equation} \label{eq:approx_likelihood}
  l_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) \mid \theta), \quad \text{where  } \epsilon > 0.
\end{equation}

and, in turn, to the approximate posterior

\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data) \propto l_{d, \epsilon}(\thetab) p(\thetab).
\end{equation}

The approximation in~\eqref{eq:approx_likelihood} is by no means the
only strategy to deal with the intractability of the likelihood
function in~\eqref{eq:likelihood}. Other strategies include modeling
the (stochastic) relationship between \(\thetab\) and \(\yb\), and its
reverse, or framing likelihood-free inference as a ratio estimation
problem, see for example \citet{blum2010, Wood2006, Papamakarios2016,
  Papamakarios2019, Chen2019, Thomas2020, Hermans2020}. However, both
OMC and robust OMC, which we introduce next, are based on the
approximation in~\eqref{eq:approx_likelihood}.

\subsection{Optimization Monte Carlo (OMC)}
\input{latex_files/ch2/2-2-OMC.tex}

\subsection{Robust optimization Monte Carlo (ROMC)}
\input{latex_files/ch2/2-3-ROMC.tex}

>>>>

An implicit or simulator-based model is a parameterized stochastic
data generating mechanism. The key characteristic of simulator-based
models is that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterized
family of probability density functions
\(\{ p(\yb|\thetab)\}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these scenarios, we can only a
access the simulator \( m_r(\thetab) \), i.e., the black-box mechanism
(computer code) that generates samples \(\yb\) in a stochastic manner
from a set of parameters \(\thetab\). We denote the process of
obtaining samples from the simulator with
\( m_r(\thetab) \rightarrow \yb \). As shown by~\citet{Meeds2015}, it
is feasible to isolate the randomness of the simulator by introducing
a set of nuisance random variables denoted by \(\ub \sim
p(\ub)\). Therefore, for a specific tuple \((\thetab, \ub)\) the
simulator becomes a deterministic mapping \(g\), such that
\(\yb=\simulator(\thetab,\ub)\). In terms of computer code, the
stochasticity of a random process is governed by the global
seed. Although each software may handle the randomness in slightly
different ways\footnote{For example, at
  \pkg{Numpy}~\citet{harris2020array}, the pseudo-random number
  generation is based on a global state, whereas, in
  \pkg{JAX}~\citet{jax2018github} random functions consume a key that
  is passed as parameter. }, in all cases, freezing the initial seed
to a specific integer converts the simulation to a deterministic piece
of code.

Simulator-based models provide modeling freedom, including hidden
(unobserved) random variables or rule-based decisions. In fact, any
physical process that can be conceptualized as a computer program of
finite steps can be modeled as a simulator-based model. Hence,
implicit models are often used to model physical phenomena in the
natural sciences such as, e.g., genetics, epidemiology or
neuroscience. Further background on simulator-based models and example
applications can be found in the articles by~\citet{gutmann2016,
  lintusaari2017, sisson2018, cranmer2020}.

The modeling freedom of simulator-based models, however, comes at the
price of difficulties in inferring the parameters of
interest. Denoting the observed data as \(\data\), the main difficulty
lies at the intractability of the likelihood function
\(L(\thetab) = p(\data|\thetab)\). To better see the sources of the
intractability, and to address them, we go back to the basic
characterization of the likelihood as the (rescaled) probability of
generating data \(\yb\) that is similar to the observed data
\(\data\), using parameters \(\thetab\). Formally, the likelihood
\(L(\thetab)\) equals

\begin{equation}
  \label{eq:likelihood}
  L(\thetab) = \lim_{\epsilon \to 0} c_\epsilon \int_{\yb \in B_{d,\epsilon}(\data)} p(\yb|\thetab)d\yb =
  \lim_{\epsilon \to 0} c_\epsilon \Pr(\simulator(\thetab, \ub) \in \region(\data)  \mid \theta)
\end{equation}
where \(c_\epsilon\) is a proportionality factor that depends on
\(\epsilon\) and \(\region(\data)\) is an \(\epsilon\) region around
\(\data\) that is defined through a distance function \(d\), i.e., \
\(\region(\data) := \{\yb: d(\yb, \data) \leq \epsilon \}\).

Equation~\ref{eq:likelihood} highlights the main source of
intractability; computing
\(\Pr(\simulator(\thetab,\ub) \in \region(\data)) | \thetab\) as the
fraction of samples that lie inside the \(\epsilon\) region around
\(\data\) is computationally infeasible in the limit of
\(\epsilon \to 0\). Hence, the constrainted is relaxed to
\(\epsilon > 0\), which leads to the approximate likelihood:

\begin{equation}
  \label{eq:approx_likelihood}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) \mid \theta), \quad \text{where  } \epsilon > 0.
\end{equation}

and, in turn, to the approximate posterior:

\begin{equation} \label{eq:approx_posterior}
  p_{d,\epsilon}(\thetab|\data)
  \propto L_{d, \epsilon}(\thetab) p(\thetab)
  \propto L_{d, \epsilon}(\thetab) p(\thetab)
\end{equation}

The approximation in
Equations~\ref{eq:approx_likelihood},~\ref{eq:approx_posterior} is by
no means the only strategy to deal with the intractability of the
likelihood function in Equation~\ref{eq:likelihood}. Other strategies include
modeling the (stochastic) relationship between \(\thetab\) and
\(\yb\), and its reverse, or framing likelihood-free inference as a
ratio estimation problem, see for example \citet{blum2010, Wood2006,
  Papamakarios2016, Papamakarios2019, Chen2019, Thomas2020,
  Hermans2020}. However, both OMC and robust OMC, which we introduce
next, are based on the approximation in Equation~\ref{eq:approx_likelihood}.

\subsection{Optimization Monte Carlo (OMC)}

Our description of OMC~\citet{Meeds2015}
follows~\citet{Ikonomov2019}. We define the indicator function (boxcar
kernel) that equals one only if \(\xb\) lies in \(\region(\yb)\):
%
\begin{equation}
  \label{eq:indicator}
  \indicator{\region(\yb)}(\xb)=
  \left\{
    \begin{array}{ll}
      1 & \mbox{if } \xb \in \region(\yb) \\
      0 & \mbox{otherwise}
    \end{array} \right. \end{equation}
%

We, then, rewrite the approximate likelihood function
\(L_{d, \epsilon}(\thetab)\) of Equation~\ref{eq:approx_likelihood} as:

\begin{gather}
  \label{eq:approx_likelihood_omc}
  L_{d, \epsilon}(\thetab) = \Pr(\yb \in \region(\data) | \thetab) =
  \int_{\ub} \indicator{\region(\data)}(g(\thetab, \ub)) d \ub
\end{gather}

which can be approximated using samples from the simulator:

\begin{equation}
  \label{eq:samples_approx_likelihood_omc}
  L_{d, \epsilon}(\thetab) \approx \frac{1}{N} \sum_{i=1}^N \indicator{\region (\data)} (\simulator(\thetab, \ub_i))
 \quad \text{ where } \ub_i \sim p(\ub).
\end{equation}

In Equation~\ref{eq:samples_approx_likelihood_omc}, for each
\(\ub_i\), there is a region \(\accregioni\) in the parameter space
\(\thetab\) where the indicator function returns one, i.e.,
\(\accregioni = \{ \thetab: \simulator(\thetab, \ub_i) \in
\region(\data) \}\). Therefore, we can approximate the likelihood as:

\begin{equation}
  \label{eq:alt_view_theta}
  L_{d, \epsilon}(\thetab) \approx \frac{1}{N} \sum_{i=1}^N \indicator{\accregioni}(\thetab)
\end{equation}

and the corresponding approximate posterior:

\begin{equation}
  \label{eq:approx_posterior_omc}
  p_{d,\epsilon}(\thetab|\data) \propto
  p(\thetab) \sum_i^N  \indicator{\accregioni}(\thetab).
\end{equation}

As argued by \citet{Ikonomov2019}, these derivations provide a unique
perspective for likelihood-free inference by shifting the focus onto
the geometry of the acceptance regions \(\accregioni\). Indeed, the
task of approximating the likelihood and the posterior becomes a task
of characterising the sets \(\accregioni\). OMC by \citet{Meeds2015}
assumes that the distance \(d\) is the Euclidean distance
\(||\cdot||_2\) between summary statistics \(\Phi\) of the observed
and generated data, and that the \(\accregioni\) can be well
approximated by infinitesimally small ellipses. These assumptions lead
to an approximation of the posterior in terms of weighted samples
\(\thetab_i^*\) that achieve the smallest distance between observed
and simulated data for each realisation \(\ub_i \sim p(\ub)\), i.e.,\
\begin{equation} \label{eq:omc_opt_prob}
\thetab_i^* = \argmin_{\thetab} ||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2  , \quad \ub_i \sim p(\ub).
\end{equation}
The weighting for each \(\thetab_i^*\) is proportional to the prior
density at \(\thetab_i^*\) and inversely proportional to the determinant
of the Jacobian matrix of the summary statistics at \(\thetab_i^*\). For
further details on OMC we refer the reader to \citep{Meeds2015,
  Ikonomov2019}.

\subsection{Robust optimization Monte Carlo (ROMC)}

\citet{Ikonomov2019} showed that considering infinitesimally small
ellipses can lead to highly overconfident posteriors. We refer the
reader to their paper for the technical details and conditions for
this issue to occur. Intuitively, it happens because the weights in
OMC are only computed from information at \(\thetab_i^*\), and using
only local information can be misleading. For example if the curvature
of \(||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2\) at
\(\thetab_i^*\) is nearly flat, the curvature alone may wrongly
indicate that \(\accregioni\) is much larger than it actually is. In
our software package we implement the robust generalisation of OMC by
\citet{Ikonomov2019} that resolves this issue.

ROMC, firstly, approximates the acceptance regions \(\accregioni\),
and defines proposal distributions \(q_i(\thetab)\) on them. The
proposal distributions are used for generating posterior samples
\(\thetab_{ij} \sim q_i\). The samples are assigned (importance)
weights \(w_{ij}\) that compensate for using the proposal
distributions \(q_i(\thetab)\) and not the prior \(p(\thetab)\),
\begin{equation}
  w_{ij} = \frac{\indicator{\accregioni}(\thetab_{ij}) p(\thetab_{ij})}{q(\thetab_{ij})}.
  \label{eq:sampling}
\end{equation}
Given the weighted samples, any expectation
\(\E_{p(\thetab|\data)}[h(\thetab)]\) of some function \(h(\thetab)\), can be approximated as
\begin{equation} \label{eq:expectation}
  \E_{p(\thetab|\data)}[h(\thetab)] \approx \frac{\sum_{ij} w_{ij} h(\thetab_{ij})}{\sum_{ij} w_{ij}}
\end{equation}
\citet{Ikonomov2019} considered uniform distributions as proposal
distributions so that the main task is to approximate the acceptance
regions \(\accregioni\) and to represent them so that uniform sampling
is easy. The approximation of the acceptance regions contains two
compulsory and one optional step: (1) solving the optimisation
problems as in OMC, (2) constructing bounding boxes around
\(\accregioni\) and optionally, (3) refining the approximation via a
surrogate model of the distance.

\subsubsection*{Solving the deterministic optimisation problems}
For each set of nuisance variables \(\ub_i, i = \{1,2,\ldots,n_1 \}\),
we search for a point \(\thetab^*_i\) such that
\(d(\simulator(\thetab^*_i,\ub_i), \data) \le \epsilon\). In general,
\(d\) can be any valid distance, but for the rest of the paper we
consider \(d\) as the squared Euclidean distance. For notational
convenience, we denote \(d(\simulator(\thetab,\ub_i), \data)\) as
\(d_i(\thetab)\).  Obtaining \(\theta_i^*\) involves solving the
following optimisation problem:
\begin{subequations}
\begin{alignat}{2}
  &\!\min_{\thetab}        && d_i(\thetab) \label{eq:optProb}\\
  &\text{subject to} & \quad& d_i(\thetab) \leq \epsilon
\end{alignat}
\end{subequations}
%
The optimisation problem can be treated as unconstrained, accepting
the optimal point
\(\thetab_i^* = \text{argmin}_{\thetab} d_i(\thetab)\) only if
\(d_i(\thetab_i^*) \le \epsilon\). If \(d_i(\thetab)\) is
differentiable any gradient-based optimizer can be used
for~\ref{eq:optProb}. The gradients \(\nabla_{\thetab} d_i(\thetab)\)
can be either provided in closed form or approximated by finite
differences. In case \(d_i\) is not differentiable, Bayesian
Optimisation~\citep{Shahriari2016} provides an alternative
approach. In this scenario, apart from obtaining an optimal
\(\thetab_i^* \), a surrogate model \(\hat{d}_i(\thetab)\) of the
distance function \(d_i(\thetab)\) is also automatically obtained;
\(\hat{d}_i\) can then substitute the actual distance function in
downstream steps of the algorithms, with possible computational gains
especially if evaluating the actual distance \(d_i(\thetab)\) is
expensive.

\subsubsection*{Estimating the acceptance regions}
The acceptance region \(\accregioni\) is approximated by a bounding
box \(\accregionihat\). Ideally, we want the bounding box to be as
tight as possible to \(\accregioni\) to ensure high acceptance rate in
the importance sampling, but big enough for not discarding valid
parts. The bounding boxes are built in two steps. First, we define
their axes \(\mathbf{v}_m\), \(m = \{1, \ldots, D\}\) based on the
(estimated) curvature of the distance at \(\thetab_i^*\). Second, we
determine the size of the box via a one-dimensional line-search method
along each axis, see Algorithm~\ref{alg:region_construction} for the
details. After the bounding boxes construction, a uniform distribution
\(q_i\) is defined on each bounding box, and is used as the proposal
region for importance sampling.

\subsubsection*{Refining the estimate via a local surrogate model (optional)}
When computing the weight \(w_{ij}\) in \eqref{eq:sampling}, we need
to check whether the samples \(\thetab_{ij} \sim q_i\) lie inside the
acceptance region \(\accregioni\). This can be considered to be a
safety-mechanism that corrects for any inaccuracies in the
construction of \(\accregionihat\) above. However, this check involves
evaluting the distance function \(d_i(\thetab_{ij})\), which can be
expensive if the model is complex. \citet{Ikonomov2019} thus proposed
to fit a surrogate model \(\tilde{d}_i(\thetab)\) of the distance
function \(d_i(\thetab)\), on data points that lie inside
\(\accregionihat\). They used a simple quadratic model whilst other
regression models are, in principle, possible too. The advantage of
using a quadratic model is that it has ellipsoidal isocontours, which
thus naturally allowed \citet{Ikonomov2019} to replace the bounding
box approximation of \(\accregioni\) with a tigher-fitting ellipsoidal
approximation.\footnote{The difference to the infinitesimal
  ellipsoidal model in OMC is the estimation procedure: OMC uses
  information at \(\thetab_i^*\) whilst, here, information in
  \(\accregionihat\) is used, which results in a more stable fit.}

The training data for the quadratic model is obtained by sampling
\(\thetab_{ij} \sim q_i\) and accessing the distances
\(d_i(\thetab_{ij})\). The generation of the training data adds an
extra computational cost, but leads to a significant speed-up when
evaluating the weights \(w_{ij}\). Moreover, the extra cost is largely
eliminated if Bayesian Optimisation with a Gaussian process (GP)
surrogate model \(\hat{d}_i(\thetab)\) was used to obtain
\(\thetab_i^*\) in the first step. In this case, we can use
\(\hat{d}_i(\thetab)\) instead of \(d_i(\thetab)\) to generate the
training data. This essentially replaces the global GP model with a
simpler local quadratic model which is typically more robust.
